---
title: "Biostatistics 1 - Assignment 1"
author: "Group 3: William Fleury Dziuk, Galya Gadzhalova, Chen Ma, Afroditi-Iliana Zotou"
date: "`r Sys.Date()`"
output:
  pdf_document:
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r, echo=FALSE}
#install.packages("car")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", message = FALSE, warning = FALSE)
library(tableone)
library(knitr)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
pacman::p_load(tidyverse,rms,haven,mgcv,epitools,logistf,nlpred,geepack,
skimr,pROC,tableone,emmeans,glmtoolbox,CalibrationCurves,mice, dcurves)
```

### Introduction

Here we study the outcome coronary heart disease (CHD, variable name in
our <data:chd69>). Our task is to find a prediction model for the risk
of CHD. We will consider the following covariates/predictors which are
previously known risk factors for heart disease: behaviour type A/B,
age, systolic bp, cholesterol, bmi, smoking, and arcus.

We have access to the dataset wcgs or the Western Collaborative Group
Study data. This is a prospective cohort study, that recruited
middle-aged men (ages 39 to 59) who were employees of 10 California
companies and collected data on 3154 individuals during the years
1960-1961. These subjects were primarily selected to study the
relationship between behavior pattern and the risk of coronary hearth
disease (CHD). A number of other risk factors were also measured to
provide the best possible assessment of the CHD risk associated with
behavior type. Additional variables collected include age, height,
weight, systolic blood pressure, diastolic blood pressure, cholesterol,
smoking, and corneal arcus

### Preparing for analysis and cleaning the data {.unnumbered}

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# read in the data
data(wcgs)
# create factors from var ditpat0 with levels B and A in stead of 0 and 1
wcgs$dibpat0f<-factor(wcgs$dibpat0,levels=0:1,label=c("B","A"))
# create categories for age
wcgs$agegroup <- cut(wcgs$age0,breaks=c(39,45,55,60),include.lowest = T,right = FALSE)
# binary variable for smoker or not
wcgs$smoker <- ifelse(wcgs$ncigs0>0,1,0)
wcgs$smokerf <- factor(wcgs$smoker,levels=c(0,1),labels=c("No","Yes"))
# convert height from inches to cm
wcgs$heightcm <- wcgs$height0*2.54
wcgs$weightkg <- wcgs$weight0*0.45359237
wcgs$bmi <- wcgs$weightkg / (wcgs$heightcm/100)^2
wcgs$bmicat <- cut(wcgs$bmi,breaks=c(0,25,30,40),include.lowest = T,right = FALSE)
wcgs$cholmmol <- wcgs$chol0/39
wcgs$sbp10 <- wcgs$sbp0/10
# Create categories of sbp (systolic blood pressure). Make sure to have the lowest and highe
wcgs$sbpcat <- cut(wcgs$sbp0,breaks=c(0,140,240),include.lowest = T,right = FALSE)
# For use when tabulating the data you can create labels
wcgs$chd69f <- factor(wcgs $chd69,levels=c(0,1),labels=c("No","Yes"))
wcgs$cholmmol <- ifelse(wcgs$cholmmol<15,wcgs$cholmmol,NA)
```

In this analysis we’ll consider the following variables:

chd69 age systolic bp cholesterol bmi smoking arcus behaviour type

A new dataframe was created containing only the variables relevant for
the analysis.

```{r}
d <- wcgs %>% select(id,agegroup,age0,cholmmol,sbp10,bmi,smokerf,arcus0,dibpat0f,chd69)
```

After that, all complete cases were extracted from the cleaned dataset.

```{r}
# Complete cases of the cleaned data*
dc <- d %>% drop_na()
```

### Task 1 - Table 1

**Task:** Create table 1. Describe all available variables in the data.
We will continue with the complete case data we have now called dc

**Answer:**: The following table summarizes the characteristics of the
complete case dataset (dc) for the variables we are considering.

Continuous variables are presented as mean (Standard Deviation - SD) and
categorical variables are presented as number ($n$) (%). Our complete
case data (dc) includes $n=3139$ individuals.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
vars <- c("id", "agegroup", "age0", "cholmmol", "sbp10", "bmi",
          "smokerf", "arcus0", "dibpat0f", "chd69")

# create table 1
table1 <- CreateTableOne(vars = vars, data = dc)

print(table1)
```

### Task 2 - Overall risk or overall rate

**Task:** What is the outcome we are interested in?

**Answer:** We are interested in the outcome coronary heart disease
(CHD). In our dataset the variable for this is chd69. Overall, our task
is to find a prediction model for the risk of CHD.

**Task:** What are the known risk factors for our outcome of interest?

**Answer:** The known risk factors (covariates/predictors) for our
outcome of interest, Coronary Heart Disease (CHD), are: Behaviour type
(Type A/B), Age, Systolic blood pressure (systolic bp), Cholesterol,
BMI, Smoking, Arcus (corneal arcus)

These are the variables that have been previously known as risk factors
for heart disease and are taken into consideration for our model.

*The known factors that cause risk for the particular disease are
separated into two groups (modifiable and non-modifiable)
[@cdc_heart_disease_risk_factors].*

*Modifiable risk factors are:*

-   *smoking*
-   *physical inactivity*
-   *unhealthy diet*
-   *high blood pressure*
-   *unhealthy blood cholesterol levels*

*Non-modifiable risk factors are:*

-   *sex*
-   *ethnicity*
-   *age*
-   *family history (genetics)*

**Task:** How many people are included?

**Answer:** The dataset used for this analysis contains 3.139
observations.

**Task:** What is the overall prevalence of the disease in our cohort?

**Answer:** To find the overall prevalence (the percentage of people who
have the disease), we look at the mean value of the binary outcome
variable (chd69) reported in Table 1. So, the overall prevalence is 8%

### Task 3 - Building the model and choosing predictors

#### 3a.

**Task:** Use the available data to create the optimal prediction model
for our outcome, choose the appropriate model and the predictors that
will improve your predictions. Please explain the reasons for choosing
the model type you chose and why the predictors were chosen.

**Answer:**

Before building a predictive model, a correlation matrix was calculated
to check for variables with strong correlation. The threshhold for
strong correlation is 0.8.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
numeric_vars <- dc %>% 
  select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "complete.obs")
round(cor_matrix, 2)

```

**Table 1: Correlation Matrix of Numeric Variables**

|          | id    | age0  | cholmmol | sbp10 | bmi   | arcus0 | chd69 |
|----------|-------|-------|----------|-------|-------|--------|-------|
| id       | 1.00  | -0.05 | 0.06     | -0.05 | 0.03  | -0.02  | 0.00  |
| age0     | -0.05 | 1.00  | 0.09     | 0.17  | 0.03  | 0.19   | 0.12  |
| cholmmol | 0.06  | 0.09  | 1.00     | 0.12  | 0.07  | 0.13   | 0.15  |
| sbp10    | -0.05 | 0.17  | 0.12     | 1.00  | 0.29  | 0.03   | 0.13  |
| bmi      | 0.03  | 0.03  | 0.07     | 0.29  | 1.00  | -0.03  | 0.06  |
| arcus0   | -0.02 | 0.19  | 0.13     | 0.03  | -0.03 | 0.19   | 0.07  |
| chd69    | 0.00  | 0.12  | 0.15     | 0.13  | 0.06  | 0.07   | 1.00  |

The matrix shows that none of the variables exceed the threshold of 0.8,
therefore there is no risk for multicollinearity in this dataset.
Meaning that it is not necessary to remove any of the variables at this
stage of the analysis, since every variable contributes uniquely.

A logistic regression model was selected for this analysis because the
outcome variable is binary and diagnostic checks indicated no evidence
of multicollinearity among the predictors.

The next step was to fit the full logistic regression model which
included all available predictors in the cleaned dataset. This model was
then compared to a reduced model obtained using backward stepwise
selection, where predictors were removed based on their contribution to
the model fit. In our case the function only excluded one of the initial
variables (agegroup).

The result yelds that the step model had better performance compared to
the original one. This is concluded since its AIC value is lower (namely
1594.23) than in the full model (namely AIC of 1597.07). This result is
also conceptually reasonable since agegroup and age0 represent the same
underlying information and including both of them in the model introduces
redundancy. Moreover, age0 is more informative as a continuous variable
and agegroup was appropriately removed by the stepwise algorithm.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
full_model <- glm(chd69 ~  age0 + cholmmol + sbp10 + bmi + smokerf +
                    arcus0 + dibpat0f + agegroup,
                  data = dc,
                  family = binomial)


step_model <- step(full_model, direction = "backward")

```

To assess whether the continuous predictors satisfy the logistic
regression assumption of linearity in the logit, scatterplots of each
predictor against the predicted log-odds were created. They show that
the requirement for linearity is met and step-wise logistic regression
could be used.

```{r, echo=FALSE}


dc$logit <- predict(step_model, type = "link")

# Arrange plots in 2 x 2 grid
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Age
plot(dc$age0, dc$logit, main = "Age vs Logit")
lines(lowess(dc$age0, dc$logit), col = "red", lwd = 2)

# Cholesterol
plot(dc$cholmmol, dc$logit, main = "Cholesterol vs Logit")
lines(lowess(dc$cholmmol, dc$logit), col = "red", lwd = 2)

# Systolic BP
plot(dc$sbp10, dc$logit, main = "SBP10 vs Logit")
lines(lowess(dc$sbp10, dc$logit), col = "red", lwd = 2)

# BMI
plot(dc$bmi, dc$logit, main = "BMI vs Logit")
lines(lowess(dc$bmi, dc$logit), col = "red", lwd = 2)

```

In the following table the coefficients of the step-wise model are
presented. Only one of them is not statistically significant.

```{r, echo=FALSE}
coef_table <- data.frame(
  Variable = c("(Intercept)", "age0", "cholmmol", "sbp10", "bmi", 
               "smokerfYes", "arcus0", "dibpat0fA"),
  Estimate = c(-12.24944, 0.05725, 0.40142, 0.18205, 0.05928,
               0.60023, 0.22038, 0.70145),
  Std_Error = c(0.98648, 0.01226, 0.06004, 0.04156, 0.02659,
                0.14191, 0.14340, 0.14510),
  Z_value = c(-12.417, 4.669, 6.686, 4.380, 2.229,
              4.230, 1.537, 4.834),
  P_value = c("< 2e-16", "3.03e-06", "2.30e-11", "1.18e-05", "0.0258",
              "2.34e-05", "0.1243", "1.34e-06")
)

kable(coef_table, digits = 4, caption = "Logistic Regression Coefficient Table")
```

Therefore, the chosen step-wise model has the following form:

$$
\begin{aligned}
\text{logit}\big(P(\text{CHD}_i = 1)\big)
&= \beta_0
+ \beta_1 \cdot \text{age0}_i
+ \beta_2 \cdot \text{cholmmol}_i
+ \beta_3 \cdot \text{sbp10}_i \\
&\quad
+ \beta_4 \cdot \text{bmi}_i
+ \beta_5 \cdot \text{smoker}_i
+ \beta_6 \cdot \text{arcus0}_i
+ \beta_7 \cdot \text{dibpat0fA}_i.
\end{aligned}
$$

Below are presented the AIC (Akaike Information Criterion) and BIC
(Bayesian Information Criterion) values of the full model and the
stepwise model. In general, the lower the AIC and BIC values, the more
optimal the model is. Both criteria favored the stepwise model, which
indicates that the stepwise model provides a better balance between
goodness of fit and model complexity.

**Table: Comparison of AIC and BIC Between Models**

| Metric               | Value    |
|----------------------|----------|
| AIC (Full Model)     | 1597.072 |
| AIC (Stepwise Model) | 1594.227 |
| BIC (Full Model)     | 1657.589 |
| BIC (Stepwise Model) | 1642.640 |

#### 3b.

**Task:** Are there any predictors that should be included as
interaction terms with other variables or as categorical variables? If
so, please explain the reasoning for this in the model?

**Answer:**

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

interaction_model <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    age0:cholmmol +
    age0:sbp10 +
    smokerf:cholmmol +
    smokerf:sbp10,
  data = dc,
  family = binomial
)


interaction_model_smookerf_sbp10 <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    smokerf:sbp10,
  data = dc,
  family = binomial
)


interaction_model_smoker_cholmmol <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    smokerf:cholmmol,
  data = dc,
  family = binomial
)


interaction_model_age0_sbp <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    age0:sbp10,
  data = dc,
  family = binomial
)


interaction_model_age0_cholmmol <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    age0:cholmmol,
  data = dc,
  family = binomial
)


interaction_model_bmi_cholmmol <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    bmi:cholmmol,
  data = dc,
  family = binomial
)

AIC(step_model, interaction_model)
BIC(step_model, interaction_model)

AIC(step_model, interaction_model_age0_cholmmol)
BIC(step_model, interaction_model_age0_cholmmol)

AIC(step_model, interaction_model_age0_sbp)
BIC(step_model, interaction_model_age0_sbp)

AIC(step_model, interaction_model_smoker_cholmmol)
BIC(step_model, interaction_model_smoker_cholmmol)

AIC(step_model, interaction_model_smookerf_sbp10)
BIC(step_model, interaction_model_smookerf_sbp10)

AIC(step_model, interaction_model_bmi_cholmmol)
BIC(step_model, interaction_model_bmi_cholmmol)


summary(interaction_model)
summary(interaction_model_age0_cholmmol)
summary(interaction_model_age0_sbp)
summary(interaction_model_smoker_cholmmol)
summary(interaction_model_smookerf_sbp10)
summary(interaction_model_bmi_cholmmol)

interaction_model$aic
interaction_model_age0_cholmmol$aic
interaction_model_age0_sbp$aic
interaction_model_bmi_cholmmol$aic
interaction_model_smoker_cholmmol$aic
interaction_model_smookerf_sbp10$aic

```

Several interaction terms were tested to assess whether the effect of
one predictor on CHD depended on another predictor. We mainly
investigated two-way interactions, since three- or four-way interactions
would be difficult to interpret clinically and would potentially lead to
overfitting.

The interactions were selected based on clinical reasoning and included
age $\times$ cholesterol, age $\times$ systolic blood pressure, smoking $\times$
cholesterol, smoking $\times$ systolic blood pressure, and BMI $\times$
cholesterol.

First, a model containing all interaction terms was fitted. Then,
individual models were fitted in which each interaction term was added
separately together with all main predictors. For each model, AIC values
were compared with those from the stepwise model and the significance of
the interaction term was evaluated. Across these models, most
interaction terms did not improve model performance and were not
statistically significant.

Only the BMI $\times$ cholesterol interaction improved the model by lowering
the AIC and showing statistical significance.This model achieved the
lowest AIC ($1588.306$) and the lowest BIC ($1642.771$). This indicates
that, while most predictors act independently in this dataset, the BMI
$\times$ cholesterol interaction slightly modifies the effect of one
predictor on CHD. The above can be verified from the following table.

```{r, echo=FALSE}
model_list <- list(
  "1. Stepwise Model (Base)" = step_model,
  "2. Age x Cholesterol" = interaction_model_age0_cholmmol,
  "3. Age x SBP" = interaction_model_age0_sbp,
  "4. Smoker x Cholesterol" = interaction_model_smoker_cholmmol,
  "5. Smoker x SBP" = interaction_model_smookerf_sbp10,
  "6. BMI x Cholesterol" = interaction_model_bmi_cholmmol
)
# Extract AIC and BIC values for all models
aic_values <- sapply(model_list, AIC)
bic_values <- sapply(model_list, BIC)

# Combine the results into a data frame
comparison_table <- data.frame(
  AIC = aic_values,
  BIC = bic_values
)

# Optional: Use a package like 'knitr' or 'flextable' to render a clean output
print(comparison_table)
```

Furthermore, a likelihood ratio test was conducted to compare the
stepwise-selected main‐effects model with the model that included the
BMI $\times$ cholesterol interaction term.

```{r lrt-test, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
anova(step_model, interaction_model_bmi_cholmmol, test = "LRT")
```

```{r, echo=FALSE}
library(knitr)

lrt_table <- data.frame(
  Model = c("Stepwise Model (no interaction)", "Interaction Model (with bmi:cholmmol)"),
  Residual_Df = c(3131, 3130),
  Residual_Deviance = c(1578.2, 1570.3),
  Df = c("", 1),
  Deviance = c("", 7.9208),
  P_value = c("", "0.004887")
)

kable(lrt_table, 
      caption = "Likelihood Ratio Test Comparing Models With and Without the BMI × Cholesterol Interaction",
      align = "c")

```

The p-value in the table represents the deviance difference. A high
p-value means smaller chi-square statistic and hence smaller deviance
difference $\Delta D$. In our case, the p-value $= 0.0049<0.01$
indicates that this reduction is statistically significant. This result
suggests that the effect of cholesterol on CHD risk depends on BMI, and
that including the bmi × cholmmol interaction provides additional
information and significantly improves the model fit.

After the aforementioned tests, the final model that is chosen is:

$$\text{CHD} \sim \text{Age} + \text{Cholesterol} + \text{Systolic BP} + \text{BMI} + \text{Smoking Status} + \text{Corneal Arcus} + \text{Behavior Type} + \text{Cholesterol} : \text{BMI}$$
and is mathematically expressed as:

$$
\begin{aligned}
\text{logit}\big(P(\text{CHD}_i = 1)\big)
&= \beta_0 
+ \beta_1 \cdot \text{age0}_i
+ \beta_2 \cdot \text{cholmmol}_i
+ \beta_3 \cdot \text{sbp10}_i \\
&\quad
+ \beta_4 \cdot \text{bmi}_i
+ \beta_5 \cdot \text{smoker}_i
+ \beta_6 \cdot \text{arcus0}_i \\
&\quad
+ \beta_7 \cdot \text{dibpat0fA}_i
+ \beta_8 \cdot (\text{cholmmol}_i \cdot \text{bmi}_i).
\end{aligned}
$$

```{r, echo=FALSE}
coef_table_interaction <- data.frame(
  Variable = c("(Intercept)", "age0", "cholmmol", "sbp10", "bmi",
               "smokerfYes", "arcus0", "dibpat0fA", "cholmmol:bmi"),
  Estimate = c(-21.28332, 0.05645, 1.87548, 0.18684, 0.41952,
               0.59778, 0.21903, 0.72663, -0.05908),
  Std_Error = c(3.30528, 0.01228, 0.51871, 0.04159, 0.12774,
                0.14197, 0.14380, 0.14564, 0.02063),
  Z_value = c(-6.439, 4.595, 3.616, 4.492, 3.284,
              4.211, 1.523, 4.989, -2.864),
  P_value = c("1.20e-10", "4.32e-06", "0.00030", "7.05e-06", "0.00102",
              "2.55e-05", "0.12771", "6.07e-07", "0.00418")
)

kable(coef_table_interaction, digits = 4,
      caption = " Logistic Regression Coefficients for the Model Including the BMI × Cholesterol Interaction")
```

From Table 5, where the numerical estimates, errors, and significance
tests for each $\beta$ are presented, we can also verify that the
interaction term (cholmmol:bmi) is statistically significant.

#### 3c.

**Task:** Calculate the predicted risk of the outcome for every person
in the dataset according to the model you have chosen and add it to your
dataset.

**Answer:**

We calculated the predicted risk of CHD for each individual in the
dataset using the chosen interaction model ($bmi \times cholmmol$) and
stored these values in a new column called predicted_risk. The table
below shows the top rows of the final dataset:

```{r}
dc$predicted_risk <- predict(interaction_model_bmi_cholmmol, newdata = dc, type = "response")
kable(head(dc), digits=3)
```

### Task 4 - Discrimination

#### 4a.

**Task:** To discriminate between the cases and non-cases please plot
the ROC curve and calculate the derived AUC including 95% confidence
intervals.

**Answer:**

According to Dr. Thorgerdur Palsdottir presentation notes, the ROC Curve (Receiver operating curve) and and the AUC (Area Under the Curve) evaluate how well the predicted risk from our model distinguishes between patients with and without the disease.

```{r}
library(pROC)
#ROC curve object using pROC package
roc_curve <- roc(dc$chd69, dc$predicted_risk)

#Plotting ROC Curve
plot(roc_curve, col = "blue", main = "ROC Curve", print.auc = TRUE)

#AUC
auc_app <- auc(roc_curve)

# 95% CI for AUC
auc_ci <- ci.auc(roc_curve)

```

The derived AUC (area under the curve) is $0.753$ which indicates good discrimination.
The $95\%$ Confidence Interval for AUC is $(0.7237,0.7824)$.
The ROC curve plots sensitivity (true positive rate) vs 1-specificity (false positive rate) across all possible thresholds.The diagonal gray line represents random selection (AUC = 0.5).
Our curve is above this line and has a significant predictive value.

#### 4b.

**Task:** Find the threshold that maximizes the sum of the sensitivity
and specificity. Please report the sensitivity and specificity at that
threshold.

**Answer:**
In general, we define  sensitivity and specificity as:
$$\text{Sensitivity} = \frac{\text{True Positives}}{\text{True Positives} + \text{False Negatives}}$$

$$\text{Specificity} = \frac{\text{True Negatives}}{\text{True Negatives} + \text{False Positives}}$$

Our code provides the following results:
```{r, echo=FALSE}
#Make a data frame with the sensitivities, specificity, and thresholds
roc_df <- data.frame(roc_curve$sensitivities,roc_curve$specificities, roc_curve$thresholds)

#Make a column for the sum of the sensitivities and specificity
roc_df$sum <- (roc_df$roc_curve.sensitivities + roc_df$roc_curve.specificities)

#Find the threshold, sens, and spec when sum is max
roc_df[roc_df$sum == max(roc_df$sum), 1:3]

```

The threshold that maximizes the sum of the sensitivity and specificity
is 0.09042859, where sensitivity = 0.6811024 and specificity = 0.7178510. Practically this means that, at this threshold, the model correctly identifies 68.11% of the men who actually have CHD (Sensitivity) and correctly identifies 71.79% of the men who do not have CHD (Specificity).

At the optimal threshold of 9.04%, we have a good balance between the two metrics which are both around 70%. 

#### 4c.

**Task:** AUC adjusted for optimism. To adjust for optimism in the
predictions, we can use the bootstrapping method using 200 repetitions.
Please calculate the adjusted AUC with 95% confidence intervals and
compare it to the unadjusted AUC. Is there a difference in the
unadjusted and the adjusted AUC?

Note: please do not use pre-programmed functions in R to answer this
question.

**Answer:**

To adjust AUC for optimism, we used a bootstrap method with 200
repetitions, without pre-programmed functions in R. In our code, we
started with bootstrap sampling, i.e. for each iteration we took a
sample with replacement from the original dataset (boot_data). Then, we
refit the model on each bootstrap sample and calculate the AUC on the
bootstrap sample itself (boot_auc). We then applied the same fitted
model to the original dataset and calculated the AUC there (test_auc).

The difference between these two AUCs (boot_auc – test_auc) is the
optimism for that bootstrap sample and the adjusted AUC is the
difference between the apparent AUC (auc_app) and the mean optimism
(optimism_mean).

```{r}
#Set the seed for reproducability 
set.seed("206")

optimism <- numeric(200)
for (i in 1:200){
  #Sample with replacement -> bootstrap sample boot_data
  boot_index <- sample(1:nrow(dc), replace = TRUE)
  boot_data <- dc[boot_index, ]
  
  #Fit model on boot_data
  boot_model <- glm( chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf +
                       arcus0 + dibpat0f + bmi:cholmmol,
                     data = boot_data,
                     family = binomial)
  
  
  #Find AUC for boot_data
  boot_data$boot_pred <- predict(boot_model, 
                                 newdata = boot_data, 
                                 type = "response")
  boot_roc <- roc(boot_data$chd69,boot_data$boot_pred)
  boot_auc <- auc(boot_roc)
  
  #AUC_test(b) on original data
  test_pred <- predict(boot_model, 
                       newdata = dc, 
                       type = "response")
  test_roc <- roc(dc$chd69,test_pred)
  test_auc <- auc(test_roc)
  
  #Optimism(b) = AUC_boot(b) - AUC_test(b)
  optimism[i] <- boot_auc - test_auc
}

#Optimism_mean = average optimism(b)
optimism_mean <- mean(optimism)

#AUC_corrected = AUC_app – optimism_mean
auc_corrected <- auc_app - optimism_mean

#c(auc_app, auc_corrected)
```

```{r,echo=FALSE}
#auc_corrected_values <- auc_app - optimism

#c(quantile(auc_corrected_values, 0.025), quantile(auc_corrected_values, 0.975))
```

Our code provides the following results:

| Metric         | AUC       | 95% CI           |
|----------------|-----------|------------------|
| Unadjusted AUC | 0.7530452 | (0.7237, 0.7824) |
| Adjusted AUC   | 0.7461568 | (0.7168, 0.7747) |

As we can see here, the adjusted AUC is slightly less than the
unadjusted AUC. This makes sense since we are subtracting the optimism
in order to get the adjusted value. The confidence interval for the
adjusted is shifted down as well.

The difference is relatively small and the overlapping confidence intervals do not imply a significant loss in model performance. 

#### 4d.

**Task:** Cross validation. Use 10-fold cross-validation for a logistic
model to obtain the adjusted AUC and compare to the unadjusted and
adjusted AUC

**Answer:**

According to Dr. Thorgerdur Palsdottir, K-fold Cross-Validation (CV) is an internal validation technique used to estimate the model's expected prediction error. 
The 10-fold CV process starts with splitting the dataset ($\text{dc}$) into $K=10$ folds. For each fold $K$, the model was trained on the remaining $K-1$ folds. The trained model was used to predict probabilities on the held-out $K$-th fold. We then pooled all held-out predictions and computed the Cross-Validation Adjusted AUC ($\text{CV AUC}$) for all of them. 

```{r}
folds <- sample(rep(1:10, length.out = nrow(dc)))

cv_pred <- rep(NA, nrow(dc))

for (i in 1:10) {
  
  train <- which(folds != i)
  test <- which(folds == i)
  
  # Train on K-1 folds
  cv_model <- glm(chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f 
                  + bmi:cholmmol,
               data = dc[train, ], 
               family = binomial)
  
  cv_pred[test] <- predict(cv_model,
                            newdata = dc[test, ], 
                            type = "response")
}

cv_roc <- roc(dc$chd69, cv_pred)
cv_auc <- auc(cv_roc)
```

In the following table, we compare the unadjusted AUC to the two internal validation methods:

| Metric                          | Value    | Difference from Unadjusted AUC |
|---------------------------------|----------|--------------------------------|
| Unadjusted AUC                  | 0.7530   | –                              |
| Bootstrap Optimism Adjusted AUC | 0.7462   | -0.0068                        |
| Cross-Validation Adjusted AUC   | 0.7415   | -0.0115                        |

The Cross Validation adjusted AUC is less than the unadjusted AUC and is
slightly less and the Bootstrap Optimism Adjusted AUC.


Both adjusted AUCs are lower than the unadjusted AUC which means there was indeed optimism in our initial predictions. However, the differences are relatively small which means that we can consider the model stable and not overfitted to the training data.

### Task 5 - Calibration

#### 5a.

**Task:** Calibration Curve. To evaluate the performance of the
prediction model we have computed the AUC from ROC analysis, now please
plot the calibration curve and report the slope and the intercept of the
calibration curve. Use the model you chose before.

**Answer:**

According to Dr. Thorgerdur Palsdottir presentation notes, Calibration
of a Prediction Model evaluates agreement between predicted risk and
observed values in our data. We have two main tools for evaluation of
calibration: the Calibration Plot, the Calibration-in-the-large and the
Hosmer-Lemeshow test for goodness-of-fit

The calibration model is fitted by regressing the outcome (chd69) on the
linear predictor (log-odds) from our main prediction model.
$$\text{logit}(P(\text{CHD}_i=1)) = \text{Intercept} + \text{Slope} \cdot \text{logit}(\hat{P}_i)$$

```{r,echo=FALSE}

# Fit a logistic regression model, the calibration model
# Calculate the predicted log-odds (the linear predictor)
dc$predicted_logit <- predict(interaction_model_bmi_cholmmol, newdata = dc, type = "link" )

library(rms) # Requires the rms package for the lrm function

# The calibration model should be fit on the linear predictor (predicted_logit)
# without an offset if you are calculating the intercept and slope directly:
calibration_model <- lrm(chd69 ~ predicted_logit, data = dc)

# Intercept and Slope
calibration_intercept <- coef(calibration_model)[1] # Intercept 
calibration_slope <- coef(calibration_model)[2]     #Slope

#print(paste("Calibration Intercept:", calibration_intercept))
#print(paste("Calibration Slope:", calibration_slope))

# Plot the Calibration Curve (using the CalibrationCurves package)

library(CalibrationCurves)

# Note: The glm object (interaction_model_bmi_cholmmol) or the predicted probabilities
# and outcomes are used here. We'll use the predicted probabilities:

CalibrationCurve <- valProbggplot(dc$predicted_risk,dc$chd69,group = 10)

plot(CalibrationCurve)      
```

The calibration model gives an Intercept value of $2.00 \times 10^{-13}$
which is very close to 0 and means that there is no systematic bias
(predictions are not systematically too high or too low). The Slope
value which is approximately $1.00$ also indicates an excellent fit.

The calibration curve shows how well the predicted risks from our model
align with the observed risks of CHD in our cohort and compares
predicted probabilities of CHD against observed outcomes, grouped into
deciles. The diagonal line represents the perfect calibration. In the
range between $0.0 - 0.20$ for the Predicted Probability, the black
curve is very close to the ideal gray one which indicates a good
calibration for individuals predicted to have a low risk of CHD.
However, as the predicted probability increases, our model's curve falls
below the gray line which is a sign of over-prediction, i.e. our model
predicts risks that are higher than what is actually observed in the
cohort. Our curve ends at a Predicted Probability of approximately 0.60.
This means that our model is not predicting a risk higher than $60\%$
because no one in our dataset was predicted to have a risk that high.

#### 5b.

**Task:** Please estimate the goodness of fit by the method of Hosmer
and Lemeshow. Interpret the test results

**Answer:** According to Dr. Thorgerdur Palsdottir presentation
notes, the Hosmer–Lemeshow (HL) works by grouping individuals by
predicted risk, typically into 10 deciles and comparing the observed to
the expected events within each group.

```{r}
library(ResourceSelection)

# Hosmer-Lemeshow test on the final logistic regression model default: G=10 groups/deciles 
hosmer_lemeshow_result<-hoslem.test(dc$chd69, fitted(interaction_model_bmi_cholmmol), g=10)

# Print the result
print(hosmer_lemeshow_result)
```

In our case, the Hosmer-Lemeshow test provides a p-value equal to
$0.2574$ greater than the significance level 0.05. So, we
fail to reject the null hypothesis ($H_0$) and we can conclude that
there is no statistically significant difference between the observed
event rates and the predicted event rates across the 10 risk groups.
Hence, our logistic regression model has sufficient goodness of fit to
the data. This finding is consistent with the intercept and slope values
in Task 5a.

#### 5c.

**Task:** Create a prediction model only using variable agegroup as a
predictor and estimate the discrimination

**Answer:**

The prediction model using only agegroup has the following form:
$$\text{logit}(P(\text{CHD}_i=1)) = \beta_0 + \beta_1 \cdot \text{agegroup}_{[45,55)} + \beta_2 \cdot \text{agegroup}_{[55,60]}$$
The lowest age group, [39, 45) is the reference level.

```{r}
# prediction model using only agegroup
# data=dc complete case data 
model_agegroup <- glm(chd69 ~ agegroup,data = dc,family = binomial)

# predicted probabilities
dc$predicted_risk_agegroup <- predict(model_agegroup,newdata = dc,type = "response")

library(pROC)

# ROC object
roc_agegroup <- roc(dc$chd69,dc$predicted_risk_agegroup, ci = TRUE )

auc_agegroup <- roc_agegroup$auc # AUC value

# Print the AUC value and the 95% CI
print(paste("AUC for agegroup model:", auc_agegroup))
print(ci(roc_agegroup))
```

The discrimination of this model measured by the AUC is
$0.6091$ which indicates that the model has poor
discrimination, if we consider that a random guess has AUC =
$0.50$. The respective 95% confidence interval is
$(0.5765,0.6416)$

#### 5d.

**Task:** Please compare the discrimination to the model you used before
with a statistical test and interpret the results.

**Answer:**
We want to compare our final model which is:

$$\text{CHD} \sim \text{Age} + \text{Cholesterol} + \text{Systolic BP} + \text{BMI} + \text{Smoking Status} + \text{Corneal Arcus} + \text{Behavior Type} + \text{Cholesterol} : \text{BMI}$$
and we want to compare it  to the simple age-only model which is 

$$\text{CHD} \sim \text{AgeGroup}$$

We use the DeLong test for two correlated ROC curves:

```{r}
# DeLong test to compare the two AUCs
# order is important Compare the final model to the simpler model
delong_test_result <- roc.test(
  roc_curve,           # Model 1 - Our final Model
  roc_agegroup,        # Model 2 - Age-only Model
  method = "delong"
)

print(delong_test_result)
```

Our final model has a higher AUC of $0.7530$ than the Age-Only Model's
AUC which is $0.6091$. Their difference ($0.1439$) indicates a much
stronger predictive ability for our final model. Generally, the higher AUC, the better prediction the model has. Since the P-value is ($< 2.2 \times 10^{-16}$), we definitely reject the null hypothesis that the two AUCs are equal. So, the additional traditional cardiovascular
risk factors (cholesterol, SBP, BMI, smoking, arcus, and behaviour type)
leads to a statistically better model compared to a model based only on the
age group. We note that the AUCs are comparable since the predictions are made on same cohort. 

#### 5e.

**Task:** Plot both ROC curves in one figure.

**Answer:**

```{r,echo=FALSE}
library(pROC)

# Plot Final Model
# The final model has the higher AUC, so we use it to initialize the plot.
plot(roc_curve,col = "blue",             
  main = "ROC Curve Comparison: Final Model vs. Age-Only Model",
  print.auc = TRUE,            # Print AUC on the plot
  print.auc.y = 0.4,           # Position the AUC text slightly below the Age-Only AUC
  legacy.axes = TRUE,percent = TRUE
)

# Plot Age-Only Model
# Use 'lines' instead of 'plot' to add to the existing figure.
lines(roc_agegroup,
  col = "green",             # Red color for the Age-Only Model
  print.auc = TRUE,            # Print AUC on the plot
  print.auc.y = 0.35,          # Position the AUC text slightly below the Full Model AUC
  print.auc.x = 60,legacy.axes = TRUE,percent = TRUE
)


legend("bottomright",              
  legend = c("Final Model (AUC = 0.7530)","Age-Only Model (AUC = 0.6091)"),
  col = c("blue", "green"), # Corresponding colors
  lwd = 2                      # Line thickness
)

```


From the combined plot, we see a large gap between the blue curve and
the green curve which confirms that the additional traditional
cardiovascular risk factors significantly ameliorate the model fit and our final selected model
is obviously better than the simpler model.

# References
