---
title: "Biostatistics 1 - Assignment 1"
author: "Group 3: William Fleury Dziuk, Galya Gadzhalova, Chen Ma, Afroditi-Iliana Zotou"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{r}
#install.packages("car")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", message = FALSE, warning = FALSE)
library(tableone)
library(knitr)


```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
pacman::p_load(tidyverse,rms,haven,mgcv,epitools,logistf,nlpred,geepack,
skimr,pROC,tableone,emmeans,glmtoolbox,CalibrationCurves,mice, dcurves)
```

# Preparing for analysis and cleaning the data {.unnumbered}

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# read in the data
data(wcgs)
# create factors from var ditpat0 with levels B and A in stead of 0 and 1
wcgs$dibpat0f<-factor(wcgs$dibpat0,levels=0:1,label=c("B","A"))
# create categories for age
wcgs$agegroup <- cut(wcgs$age0,breaks=c(39,45,55,60),include.lowest = T,right = FALSE)
# binary variable for smoker or not
wcgs$smoker <- ifelse(wcgs$ncigs0>0,1,0)
wcgs$smokerf <- factor(wcgs$smoker,levels=c(0,1),labels=c("No","Yes"))
# convert height from inches to cm
wcgs$heightcm <- wcgs$height0*2.54
wcgs$weightkg <- wcgs$weight0*0.45359237
wcgs$bmi <- wcgs$weightkg / (wcgs$heightcm/100)^2
wcgs$bmicat <- cut(wcgs$bmi,breaks=c(0,25,30,40),include.lowest = T,right = FALSE)
wcgs$cholmmol <- wcgs$chol0/39
wcgs$sbp10 <- wcgs$sbp0/10
# Create categories of sbp (systolic blood pressure). Make sure to have the lowest and highe
wcgs$sbpcat <- cut(wcgs$sbp0,breaks=c(0,140,240),include.lowest = T,right = FALSE)
# For use when tabulating the data you can create labels
wcgs$chd69f <- factor(wcgs $chd69,levels=c(0,1),labels=c("No","Yes"))
wcgs$cholmmol <- ifelse(wcgs$cholmmol<15,wcgs$cholmmol,NA)
```

A new dataframe was created containing only the variables relevant for
the analysis.

```{r}
d <- wcgs %>% select(id,agegroup,age0,cholmmol,sbp10,bmi,smokerf,arcus0,dibpat0f,chd69)
```

After that all complete cases were extracted from the cleaned dataset.

```{r}
# Complete cases of the cleaned data*
dc <- d %>% drop_na()
```

# Task 1

```{r, echo = FALSE, message = FALSE, warning = FALSE}
vars <- c("id", "agegroup", "age0", "cholmmol", "sbp10", "bmi",
          "smokerf", "arcus0", "dibpat0f", "chd69")


# create table 1
table1 <- CreateTableOne(vars = vars, data = dc)

print(table1)

```

# Taks 2

## What is the outcome we are interested in?

We are interested in the outcome coronary heart disease

## What are the known risk factors for our outcome of interest?

The known factors that cause risk for the particular disease are
separated into two groups (modifiable and non-modifiable)
[@cdc_heart_disease_risk_factors].

Modifiable risk factors are:

-   smoking
-   physical inactivity
-   unhealthy diet
-   high blood pressure
-   unhealthy blood cholesterol levels

Non-modifiable risk factors are:

-   sex
-   ethnicity
-   age
-   family history (genetics)

## How many people are included?

The data that is used for this analysis have 3139 observations.

## What is the overall prevalence of the disease in our cohort?

To find the overall prevalence or the percentage of people who have the
disease, the mean of the binary outcome variable is calculated.

```{r}
mean(dc$chd69)
```

Meaning that the overall prevalence is 8%.

# Task 3

## Part 1

Before building a predictive model, a correlation matrix was calculated
to check for variables with strong correlation. The threshold for strong
correlation is 0.8.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
numeric_vars <- dc %>% 
  select(where(is.numeric))

cor_matrix <- cor(numeric_vars, use = "complete.obs")
round(cor_matrix, 2)

```

**Table 1: Correlation Matrix of Numeric Variables**

|          | id    | age0  | cholmmol | sbp10 | bmi   | arcus0 | chd69 |
|----------|-------|-------|----------|-------|-------|--------|-------|
| id       | 1.00  | -0.05 | 0.06     | -0.05 | 0.03  | -0.02  | 0.00  |
| age0     | -0.05 | 1.00  | 0.09     | 0.17  | 0.03  | 0.19   | 0.12  |
| cholmmol | 0.06  | 0.09  | 1.00     | 0.12  | 0.07  | 0.13   | 0.15  |
| sbp10    | -0.05 | 0.17  | 0.12     | 1.00  | 0.29  | 0.03   | 0.13  |
| bmi      | 0.03  | 0.03  | 0.07     | 0.29  | 1.00  | -0.03  | 0.06  |
| arcus0   | -0.02 | 0.19  | 0.13     | 0.03  | -0.03 | 0.19   | 0.07  |
| chd69    | 0.00  | 0.12  | 0.15     | 0.13  | 0.06  | 0.07   | 1.00  |

The matrix shows that none of the variables exceed the threshold of 0.8,
therefore there is no risk for multicollinearity in this dataset.
Meaning that it is not necessary to remove any of the variables at this
stage of the analysis, since every variable contributes uniquely.

The next step was to fit the full logistic regression model, which
included all available predictors in the cleaned dataset. This model was
then compared to a reduced model obtained using backward stepwise
selection, where predictors were removed based on their contribution to
model fit. In our case the function only excluded one of the initial
variables (agegroup). The result yelds that the step model had better
performance compared to the original one. This is concluded since AIC
value is lower than in the full model. This result is also conceptually
reasonable since agegroup and age0 represent the same underlying
information and including both in the model introduces redundancy.
Moreover, age0 is more informative as a continuous variable and agegroup
was appropriately removed by the stepwise algorithm.

```{r}
full_model <- glm(chd69 ~  age0 + cholmmol + sbp10 + bmi + smokerf +
                    arcus0 + dibpat0f + agegroup,
                  data = dc,
                  family = binomial)


step_model <- step(full_model, direction = "backward")

#summary(step_model)




```

```{r, echo=FALSE}
coef_table <- data.frame(
  Variable = c("(Intercept)", "age0", "cholmmol", "sbp10", "bmi", 
               "smokerfYes", "arcus0", "dibpat0fA"),
  Estimate = c(-12.24944, 0.05725, 0.40142, 0.18205, 0.05928,
               0.60023, 0.22038, 0.70145),
  Std_Error = c(0.98648, 0.01226, 0.06004, 0.04156, 0.02659,
                0.14191, 0.14340, 0.14510),
  Z_value = c(-12.417, 4.669, 6.686, 4.380, 2.229,
              4.230, 1.537, 4.834),
  P_value = c("< 2e-16", "3.03e-06", "2.30e-11", "1.18e-05", "0.0258",
              "2.34e-05", "0.1243", "1.34e-06")
)

kable(coef_table, digits = 4, caption = "Logistic Regression Coefficient Table")
```

In the step-model only one of the variables is not statistically
significant.

```{r, echo=FALSE}
#AIC(full_model)
#AIC(step_model)


#BIC(full_model)
#BIC(step_model)

```

**Table: Comparison of AIC and BIC Between Models**

| Metric               | Value    |
|----------------------|----------|
| AIC (Full Model)     | 1597.072 |
| AIC (Stepwise Model) | 1594.227 |
| BIC (Full Model)     | 1657.589 |
| BIC (Stepwise Model) | 1642.640 |

Furthermore, we compared the AIC (Akaike Information Criterion) and BIC
(Bayesian Information Criterion) values of the full model and the
stepwise model. Both criteria favored the stepwise model, as it achieved
lower AIC and BIC values than the full model. This indicates that the
stepwise model provides a better balance between goodness of fit and
model complexity.

## Part 2

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

interaction_model <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    age0:cholmmol +
    age0:sbp10 +
    smokerf:cholmmol +
    smokerf:sbp10,
  data = dc,
  family = binomial
)


interaction_model_smookerf_sbp10 <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    smokerf:sbp10,
  data = dc,
  family = binomial
)


interaction_model_smoker_cholmmol <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
  
    smokerf:cholmmol,
  data = dc,
  family = binomial
)



interaction_model_age0_sbp <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    age0:sbp10,
  data = dc,
  family = binomial
)


interaction_model_age0_cholmmol <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    age0:cholmmol,
  data = dc,
  family = binomial
)


interaction_model_bmi_cholmmol <- glm(
  chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f +
    bmi:cholmmol,
  data = dc,
  family = binomial
)

AIC(step_model, interaction_model)
BIC(step_model, interaction_model)

AIC(step_model, interaction_model_age0_cholmmol)
BIC(step_model, interaction_model_age0_cholmmol)

AIC(step_model, interaction_model_age0_sbp)
BIC(step_model, interaction_model_age0_sbp)

AIC(step_model, interaction_model_smoker_cholmmol)
BIC(step_model, interaction_model_smoker_cholmmol)

AIC(step_model, interaction_model_smookerf_sbp10)
BIC(step_model, interaction_model_smookerf_sbp10)

AIC(step_model, interaction_model_bmi_cholmmol)
BIC(step_model, interaction_model_bmi_cholmmol)


summary(interaction_model)
summary(interaction_model_age0_cholmmol)
summary(interaction_model_age0_sbp)
summary(interaction_model_smoker_cholmmol)
summary(interaction_model_smookerf_sbp10)
summary(interaction_model_bmi_cholmmol)

interaction_model$aic
interaction_model_age0_cholmmol$aic
interaction_model_age0_sbp$aic
interaction_model_bmi_cholmmol$aic
interaction_model_smoker_cholmmol$aic
interaction_model_smookerf_sbp10$aic

```

Several interaction terms were tested to assess whether the effect of
one predictor on CHD depended on another predictor. The interactions
were selected based on clinical reasoning and included age with
cholesterol, age with systolic blood pressure, smoking with cholesterol,
smoking with systolic blood pressure, and BMI with cholesterol. First, a
model containing all interaction terms was fitted. Then, individual
models were fitted in which each interaction term was added separately
together with all main predictors. For each model, AIC values were
compared with those from the stepwise-selected main-effects model, and
the significance of the interaction term was evaluated. Across these
models, most interaction terms did not improve model performance and
were not statistically significant. Only the BMI with cholesterol
interaction improved the model by lowering the AIC and showing
statistical significance. This indicates that, while most predictors act
independently in this dataset, the BMI with cholesterol interaction may
slightly modify the effect of one predictor on CHD.

```{r}

aic_table <- data.frame(
  Model = c("Interaction Model (bmi × cholmmol)", "Stepwise Model"),
  AIC   = c(1588.306, 1594.227)
)

kable(
  aic_table,
  digits = 3,
  caption = "Table: Comparison of AIC Between Interaction Model and Stepwise Model"
)
```

```{r, echo=FALSE}
coef_table_interaction <- data.frame(
  Variable = c("(Intercept)", "age0", "cholmmol", "sbp10", "bmi",
               "smokerfYes", "arcus0", "dibpat0fA", "cholmmol:bmi"),
  Estimate = c(-21.28332, 0.05645, 1.87548, 0.18684, 0.41952,
               0.59778, 0.21903, 0.72663, -0.05908),
  Std_Error = c(3.30528, 0.01228, 0.51871, 0.04159, 0.12774,
                0.14197, 0.14380, 0.14564, 0.02063),
  Z_value = c(-6.439, 4.595, 3.616, 4.492, 3.284,
              4.211, 1.523, 4.989, -2.864),
  P_value = c("1.20e-10", "4.32e-06", "0.00030", "7.05e-06", "0.00102",
              "2.55e-05", "0.12771", "6.07e-07", "0.00418")
)

kable(coef_table_interaction, digits = 4,
      caption = " Logistic Regression Coefficients for the Model Including the BMI × Cholesterol Interaction")
```

In the Table 5 can be seen that the interaction is statistically
significant.

Furthermore, a likelihood ratio test was conducted to compare the
stepwise-selected main‐effects model with the model that included the
BMI × cholesterol interaction term.

```{r lrt-test, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
anova(step_model, interaction_model_bmi_cholmmol, test = "LRT")
```

```{r, echo=FALSE}
library(knitr)

lrt_table <- data.frame(
  Model = c("Stepwise Model (no interaction)", "Interaction Model (with bmi:cholmmol)"),
  Residual_Df = c(3131, 3130),
  Residual_Deviance = c(1578.2, 1570.3),
  Df = c("", 1),
  Deviance = c("", 7.9208),
  P_value = c("", "0.004887")
)

kable(lrt_table, 
      caption = "Likelihood Ratio Test Comparing Models With and Without the BMI × Cholesterol Interaction",
      align = "c")

```

This result suggests that the effect of cholesterol on CHD risk depends
on BMI, and that including the bmi × cholmmol interaction provides
additional information.

## Part 3

```{r}
dc$predicted_risk <- predict(interaction_model_bmi_cholmmol, newdata = dc, type = "response")

kable(head(dc), digits=3)
```

# Task 4

## a)

To discriminate between the cases and non-cases please plot the ROC curve and calculate the derived AUC including 95% confidence intervals.

```{r}
library(pROC)
#ROC curve object using pROC package
roc_curve <- roc(dc$chd69, dc$predicted_risk)

#Plotting ROC Curve
plot(roc_curve, col = "blue", main = "ROC Curve", print.auc = TRUE)

#AUC
auc_app <- auc(roc_curve)

# 95% CI for AUC
auc_ci <- ci.auc(roc_curve)


```
The derived AUC (area under the curve): $0.753$
The $95\%$ Confidence Interval for AUC: $(0.7237,0.7824)$

## b)

Find the threshold that maximizes the sum of the sensitivity and
specificity. Please report the sensitivity and specificity at that
threshold.

```{r, echo=FALSE}
#Make a data frame with the sensitivities, specificity, and thresholds
roc_df <- data.frame(roc_curve$sensitivities,roc_curve$specificities, roc_curve$thresholds)

#Make a column for the sum of the sensitivities and specificity
roc_df$sum <- (roc_df$roc_curve.sensitivities + roc_df$roc_curve.specificities)

#Find the threshold, sens, and spec when sum is max
#roc_df[roc_df$sum == max(roc_df$sum), 1:3]

```

The threshold that maximizes the sum of the sensitivity and specificity
is 0.09042859. 
Where sensitivity = 0.6811024 and specificity = 0.7178510

## c)

AUC adjusted for optimism. To adjust for optimism in the predictions, we
can use the bootstrapping method using 200 repetitions. Please calculate
the adjusted AUC with 95% confidence intervals and compare it to the
unadjusted AUC. Is there a difference in the unadjusted and the adjusted
AUC? Note: please do not use pre-programmed functions in R to answer
this question.

```{r}
#Set the seed for reproducability 
set.seed("206")

optimism <- numeric(200)
for (i in 1:200){
  #Sample with replacement -> bootstrap sample boot_data
  boot_index <- sample(1:nrow(dc), replace = TRUE)
  boot_data <- dc[boot_index, ]
  
  #Fit model on boot_data
  boot_model <- glm( chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f + bmi:cholmmol,
                     data = boot_data,
                     family = binomial)
  
  
  #Find AUC for boot_data
  boot_data$boot_pred <- predict(boot_model, 
                                 newdata = boot_data, 
                                 type = "response")
  boot_roc <- roc(boot_data$chd69,boot_data$boot_pred)
  boot_auc <- auc(boot_roc)
  
  #AUC_test(b) on original data
  test_pred <- predict(boot_model, 
                       newdata = dc, 
                       type = "response")
  test_roc <- roc(dc$chd69,test_pred)
  test_auc <- auc(test_roc)
  
  #Optimism(b) = AUC_boot(b) - AUC_test(b)
  optimism[i] <- boot_auc - test_auc
  
  
}
#Optimism_mean = average optimism(b)
optimism_mean <- mean(optimism)

#AUC_corrected = AUC_app – optimism_mean
auc_corrected <- auc_app - optimism_mean

#c(auc_app, auc_corrected)
```

```{r,echo=FALSE}
auc_corrected_values <- auc_app - optimism

c(quantile(auc_corrected_values, 0.025), quantile(auc_corrected_values, 0.975))
```
Unadjusted AUC: $0.7530452$ 
Unadjusted CI: $(0.7237,0.7824)$

Adjusted AUC: $0.7461568$
Ajusted CI: $(0.7168202, 0.7746566)$ 

As we can see here the adjusted AUC is slightly less than the unadjusted AUC, this makes sense since we are subtracting the optimism in order to get the adjusted value. The confidence interval for the adjusted is shifted down aswell. 


## d)

Cross validation. Use 10-fold cross-validation for a logistic model to
obtain the adjusted AUC and compare to the unadjusted and adjusted AUC.

```{r}

folds <- sample(rep(1:10, length.out = nrow(dc)))

cv_pred <- rep(NA, nrow(dc))

for (i in 1:10) {
  
  train <- which(folds != i)
  test <- which(folds == i)
  
  # Train on K-1 folds
  cv_model <- glm(chd69 ~ age0 + cholmmol + sbp10 + bmi + smokerf + arcus0 + dibpat0f + bmi:cholmmol,
               data = dc[train, ], 
               family = binomial)
  
  #
  cv_pred[test] <- predict(cv_model,
                            newdata = dc[test, ], 
                            type = "response")
  
}

cv_roc <- roc(dc$chd69, cv_pred)
cv_auc <- auc(cv_roc)

```

Unadjusted AUC: $0.7530452$ 
Bootstrap Optimism Adjusted AUC: $0.7461568$
Cross Validation Adjusted AUC: $0.7415$

The Cross Validation adjusted AUC is less than the unadjusted AUC and is slightly less and the Bootstrap Optimism Adjusted AuC


# References
